If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-14 GroupPolarizationStatmod/paper/ms.tex
edited intro analysis paragraph

\date{}
\begin{document}
\maketitle
% \begin{abstract}
% \noindent 
% \emph{Group polarization} is the name for a form of consensus where
% members of a like-minded group become, on average, more extreme in their
% opinions after discussing a topic. Group polarization is important because
% it may increase social and political tensions if moderates become more
% extreme. Decades of studies have replicated detections of group
% polarization, typically using an ordinal scale to measure opinions.
% However, these studies did not account for ordinal measurements, which can
% result in spurious detections of opinion change.  Lacking original data we
% can still calculate the probability that a published detection of group
% polarization was spurious, i.e., to calculate the \emph{false detection
% rate}.  Across 54 group polarization experimental conditions from ten
% representative journal articles, we calculated false detection rates to be
% between 0.52 and 0.88, with a median of 0.75, using a generative model of
% group polarization experiments seeded with empirical data.  
% We also use our model to develop experimental designs that
% achieve an acceptable false discovery rate.  Much of group polarization
% research may be unreliable. This analysis can help change that by enabling others to avoid this
% hidden pitfall. More
% broadly this work demonstrates one important way that replication success alone
% does not imply epistemic reliability.
% \end{abstract}

@@

Our analysis estimates how the probability that a published detection of group
polarization is actually a false detection. We first identified latent opinion
parameters that yield spurious group polarization in an asymptotic $N\to\infty$
model. This gives us our first suggestion that experimental conditions may be
plausibly false. Briefly, we identified latent opinion distribution parameters
that generated spurious group polarization in 54 out of 57 experimental
conditions. In other words, 95\% of all group polarization detections are
plausibly spurious due to ceiling artifacts from the continuous latent
opinions-ordinal measurement experimental design. We then analyze the family-wise
error ($\alpha$) and false detection rate (FDR) 
for each experimental condition and aggregated 
up to the study level and across all studies and experimental conditions. 
We estimate the lower twentieth percentile of experimental conditions 
have \emph{at least} $\alpha \leq 0.16$, 
with median $\alpha \geq 0.27$ and upper eightieth percentile with $\alpha \geq
0.40$ (Table~\ref{tab:quantiles}\mt{A.  –but don't have A and B; TODO: separate
FWER and FDR tables, only use 0, 20, 50, 80, 100 for probs. Rename 0 -> min, 100 ->
max.}). For the fiftieth and eightieth percentiles this
corresponds to false discovery rates of 0.75 and 0.82, respectively. 
We emphasize that these estimates and others detailed below are lower bounds on
$\alpha$ and FDR since we are using what is likely an overgeneralized,
underspecified statistical model that does not account for all the potential
variance in experimental outcomes, which would increase error rates.
We close our Analysis by calculating which $d^*$ limits family-wise error rates to
$\alpha \leq 0.05$. This is greater than $d^* = 0.8$, classically taken to 
represent ``high'' significance, in all but six experimental conditions.

>>>>

\date{\today}
\begin{document}
\maketitle
% \begin{abstract}
% \noindent 
% \emph{Group polarization} is the name for a form of consensus where
% members of a like-minded group become, on average, more extreme in their
% opinions after discussing a topic. Group polarization is important because
% it may increase social and political tensions if moderates become more
% extreme. Decades of studies have replicated detections of group
% polarization, typically using an ordinal scale to measure opinions.
% However, these studies did not account for ordinal measurements, which can
% result in spurious detections of opinion change.  Lacking original data we
% can still calculate the probability that a published detection of group
% polarization was spurious, i.e., to calculate the \emph{false detection
% rate}.  Across 54 group polarization experimental conditions from ten
% representative journal articles, we calculated false detection rates to be
% between 0.52 and 0.88, with a median of 0.75, using a generative model of
% group polarization experiments seeded with empirical data.  
% We also use our model to develop experimental designs that
% achieve an acceptable false discovery rate.  Much of group polarization
% research may be unreliable. This analysis can help change that by enabling others to avoid this
% hidden pitfall. More
% broadly this work demonstrates one important way that replication success alone
% does not imply epistemic reliability.
% \end{abstract}

@@

We now present our determinations of whether published detections of group
polarization are plausibly spurious and calculate the best-case false detection
rate for each experimental design in the corpus. We found that 95\% of all
detections of group polarization are plausibly false, 54 out of 57. We estimate the
best-case median false detection rate to be 72\% for those 54 experimental
conditions. All but seven of the 54 designs were found to require a $d^* > 0.8$
for ``significant'' Cohen's $d$ to limit the family-wise error rate to 5\%.

% We first identified latent opinion
% parameters that yield spurious group polarization in an asymptotic $N\to\infty$
% model. This gives us our first suggestion that experimental conditions may be
% plausibly false. Briefly, we identified latent opinion distribution parameters
% that generated spurious group polarization in 54 out of 57 experimental
% conditions. In other words, 95\% of all group polarization detections are
% plausibly spurious due to ceiling artifacts from the continuous latent
% opinions-ordinal measurement experimental design. We then analyze the family-wise
% error ($\alpha$) and false detection rate (FDR) 
% for each experimental condition and aggregated 
% up to the study level and across all studies and experimental conditions. 
% We estimate the lower twentieth percentile of experimental conditions 
% have \emph{at least} $\alpha \leq 0.16$, 
% with median $\alpha \geq 0.27$ and upper eightieth percentile with $\alpha \geq
% 0.40$ (Table~\ref{tab:quantiles}\mt{A.  –but don't have A and B; TODO: separate
% FWER and FDR tables, only use 0, 20, 50, 80, 100 for probs. Rename 0 -> min, 100 ->
% max.}). For the fiftieth and eightieth percentiles this
% corresponds to false discovery rates of 0.75 and 0.82, respectively. 
% We emphasize that these estimates and others detailed below are lower bounds on
% $\alpha$ and FDR since we are using what is likely an overgeneralized,
% underspecified statistical model that does not account for all the potential
% variance in experimental outcomes, which would increase error rates.
% We close our Analysis by calculating which $d^*$ limits family-wise error rates to
% $\alpha \leq 0.05$. This is greater than $d^* = 0.8$, classically taken to 
% represent ``high'' significance, in all but six experimental conditions.
